Football Championship Ticket Options

### 2018-09-15 paper.Rmd
updates

```{r importPrep}
pacman::p_load(dplyr, ggplot2, stringr, mgcv, broom)

@@

initialData = initialData %>% 
  mutate(name = gsub("\\.|,", "", name)) %>% 
  left_join(., previousYearRank, by = c("name" = "Team")) %>% 
  group_by(name)

@@

top25InitialPrices = initialData %>% 
  slice(., 1L) %>% 
  dplyr::filter(masseyYearEnd < 26) %>% 
  select(masseyYearEnd, UL_price, LL_price, name) %>% 
  mutate(average_price = (as.numeric(UL_price) + as.numeric(LL_price)) / 2) %>% 
  tidyr::gather(key, value, -name, -masseyYearEnd) %>% 
  arrange(name) %>% 
  ggplot(., aes(name, as.numeric(value), color = key)) +
  geom_point(size = 3) +
  scale_x_discrete(limits = top25PrevNames$Team) +
  theme_minimal()

@@

@@

# Hypotheses

# Methods

## Data

@@

### Sources

Given the research questions, we sought to collect data from several different sources. Given the excellent compilation of team rankings, Massey's website was used to amass rankings.

Injury counts were pulled from Don Best's website and projected winners were pulled from XXX.

@@

### Issues and Limitations

@@

### Previous Year-end Rank and Early Prices

@@

```{r}
top25InitialPrices + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(name = "") +
  scale_x_discrete(name = "")
```

The preceding visualization provides a clear picture

@@

```{r}
completeData %>% 
  select(AP, SAG, MAS, USA, FPI) %>% 
  na.omit() %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```

Generally, strong correlations are present. For upcoming models, we will opt to use AP, MAS, and USA

## Replicating Rankings Through Option Prices

### Top 10 Trends

@@

```{r}
completeData %>% 
  mutate(Rank = as.numeric(Rank), 
         dateTime = lubridate::ymd_hms(dateTime)) %>% 
  dplyr::filter(name %in% top10Ever) %>% 
  select(dateTime, UL_price, LL_price, name) %>% 
  tidyr::gather(key, value, -name, -dateTime) %>% 
  arrange(name, dateTime) %>% 
  ggplot(., aes(dateTime, as.numeric(value), color = key, group = key)) + 
  geom_path(na.rm = TRUE, linejoin = "mitre") + 
  facet_wrap(~ name) +
  theme_minimal() +
  scale_x_datetime(name = "") +
  scale_y_continuous(name = "")
```

A quick visual inspection of every team that has been in the top 10 this year might suggest persistent high confidence for Alabama, Clemson, and Georgia (but to a slightly less degree). Not surprisingly, Ohio State and Oklahoma are starting to trend upwards; meanwhile, Notre Dame has consistently high prices (relatively speaking), but on a potential downward trend.

### Volatility

While the intervals in our option price series are not always equivalent, calculating a team's option price volatility can instructive.

```{r}
volatile %>% 
  filter(name %in% top25Ever) %>% 
  select(volLL, volUL, name) %>% 
  tidyr::gather(key = key, value = value, -name) %>% 
  ggplot(., aes(name, value, color = key)) + 
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(name = "Volatility")
```

@@

First, we can see which 25 teams have had the most price changes throughout the CFP-RSVP's existence. Note that the following table displays total price changes (includes both upper and lower level changes).

@@

ggplot(changes, aes(name, Count)) + 
  geom_point() + 
  scale_x_discrete(breaks = changes$name) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(name = "")
```

A cursory glance of the listed names should look familiar.

We can also plot the series for all of those teams which have appeared in the top 25 this season:

```{r}
allHistory %>% 
  filter(allHistory$name %in% top25Ever) %>%
  group_by(name, variable) %>% 
  mutate(obs = 1:n()) %>% 
  ggplot(., aes(obs, value, color = variable)) + 
  geom_path() +
  facet_wrap(~name) +
  theme_minimal() +
  scale_x_continuous(name = "") + 
  scale_y_continuous(name = "")
```

###

```{r}
library(lme4)

top10 = completeData %>% 
  # dplyr::filter(name %in% top10Ever) %>% 
  mutate(Rank = as.numeric(Rank), 
         dateTime = lubridate::ymd_hms(dateTime), 
         averagePrice = (as.numeric(UL_price) + as.numeric(LL_price)) / 2) %>% 
  select(name, dateTime, averagePrice, Rank, AP, SAG, MAS, USA, 
         Mean, n, computerPick, publicPick) %>% 
  mutate(day = gsub("\\s.*", "", .$dateTime)) %>% 
  group_by(name, day) %>% 
  slice(n())

mixedMod = lmer(averagePrice ~ as.numeric(USA) + as.numeric(MAS) + 
                  (1|name), data = top10)
```

@@

```{r}
predictors = completeData %>% 
  select(Mean, n, name, dateTime, masseyRecord) %>% 
  mutate(wins = as.numeric(stringr::str_extract(masseyRecord, "^[0-9]+"))) %>% 
  group_by(name) %>% 
  slice(n())

@@

modLL = gam(volLL ~ s(Mean) + wins + s(n),
            data = modDat)

# summary(modLL)

@@

knitr::kable(tidy(modLL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modLL, all.terms = TRUE)

modUL = gam(volUL ~ s(Mean) + wins + s(n),
            data = modDat)

# summary(modUL)

knitr::kable(tidy(modUL))

knitr::kable(tidy(modUL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modUL, all.terms = TRUE)
```

@@

```{r}
currentPriceModDat = completeData %>% 
  select(Mean, n, name, dateTime, masseyRecord, UL_price, LL_price) %>% 
  mutate(wins = as.numeric(stringr::str_extract(masseyRecord, "^[0-9]+"))) %>% 
  group_by(name) %>% 
  slice(n())

modLL = gam(LL_price ~ s(Mean) + wins + s(n),
            data = currentPriceModDat)

# summary(modLL)

@@

knitr::kable(tidy(modLL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modLL, all.terms = TRUE)

modUL = gam(UL_price ~ s(Mean) + wins + s(n),
            data = currentPriceModDat)

# summary(modUL)

knitr::kable(tidy(modUL))

knitr::kable(tidy(modUL, parametric = TRUE))

par(mfrow = c(2, 2))

plot.gam(modUL, all.terms = TRUE)

```

>>>>

```{r importPrep}
pacman::p_load(dplyr, ggplot2, stringr, mgcv, broom, caret)

@@

initialData = initialData %>% 
  mutate(name = gsub("\\.|,", "", name)) %>% 
  left_join(., previousYearRank, by = c("name" = "Team")) %>% 
  group_by(name) %>% 
  arrange(masseyYearEnd)

@@

top25InitialPrices = initialData %>% 
  slice(., 1L) %>% 
  dplyr::filter(masseyYearEnd < 26) %>% 
  ungroup() %>% 
  mutate(name = factor(name, levels = .$name)) %>% 
  select(masseyYearEnd, UL_price, LL_price, name) %>% 
  mutate(average_price = (as.numeric(UL_price) + as.numeric(LL_price)) / 2) %>% 
  tidyr::gather(key, value, -name, -masseyYearEnd) %>% 
  arrange(name) %>% 
  ggplot(., aes(name, as.numeric(value), color = key)) +
  geom_point(size = 3) +
  scale_x_discrete(limits = top25PrevNames$Team) +
  theme_minimal()

@@

currentRankOrder = completeData %>% 
  group_by(name) %>% 
  arrange(dateTime) %>% 
  slice(n()) %>% 
  select(name, AP) %>% 
  arrange(AP)

@@

# General Research Questions

When examining the option prices, it becomes apparent that the best teams have the highest prices; conversely teams with little chance of making the CFP have very low dollar values. Among those top teams, though, what might be driving the prices? As a preliminary pass, we are going to look at current values and volatility, and what impact wins, injuries, strength of schedule, and rankings have on those metrics. Specifically, we would anticipate that volatility has a positive relationship with wins (i.e., the more wins a team has, the more the option changes), but a negative relationship with rankings (the lower the ranking, the more the option changes). Number of injuries, while certainly challenging to a given team, might not be critical at this point within the season. Nonetheless, we would anticipate that the number of injuries will drive the option value down. We can certainly imagine a world in which strength of schedule has an impact on prices and volatility -- a nightmarish schedule might all but destroy your chances of making it to the playoffs.

# Data

@@

## Sources

Given the research questions, we sought to collect data from several different sources. Given the excellent compilation of team rankings, Ken Massey's website was used to amass rankings.

Injury counts were pulled from Don Best's website and projected winners were pulled from Odds Shark.

@@

## Issues and Limitations

@@

## Previous Year-end Rank and Early Prices

@@

```{r}
top25InitialPrices + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank())
```

The preceding visualization provides a clear picture of how early prices reflect the previous year rankings. This is not terribly surprising, given that the previous year is all of the information informing pricing.

@@

```{r}
completeData %>% 
  select(AP, SAG, MAS, USA, FPI, Mean) %>% 
  na.omit() %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```

Generally, strong correlations are present. For upcoming models, we will opt to use the mean ranking. There are a few reasons for this decision: 1) rankings like AP and USA only go up to 25 and 2) we want something that will capture all of the ratings present.

## Top 10 Trends

@@

```{r}
completeData %>% 
  mutate(Rank = as.numeric(Rank), 
         dateTime = lubridate::ymd_hms(dateTime), 
         name = factor(name, levels = currentRankOrder$name)) %>% 
  dplyr::filter(name %in% top10Ever) %>% 
  select(dateTime, UL_price, LL_price, name) %>% 
  tidyr::gather(key, value, -name, -dateTime) %>% 
  arrange(name, dateTime) %>% 
  ggplot(., aes(dateTime, as.numeric(value), color = key, group = key)) + 
  geom_path(na.rm = TRUE, linejoin = "mitre") + 
  facet_wrap(~ name) +
  theme_minimal() +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank())
```

A quick visual inspection of every team that has been in the top 10 this year might suggest persistent high confidence for Alabama, Clemson, and Georgia (but to a slightly less degree). Not surprisingly, Ohio State and Oklahoma are starting to trend upwards; meanwhile, Notre Dame has consistently high prices (relative to their real chances of actually making it to the playoffs), but on a potential downward trend.

## Volatility

While the intervals in our option price series are not always equivalent, calculating a team's option price volatility can be instructive.

```{r}
volatile %>% 
  filter(name %in% top25Ever) %>% 
  mutate(name = factor(name, levels = currentRankOrder$name)) %>% 
  select(volLL, volUL, name) %>% 
  tidyr::gather(key = key, value = value, -name) %>% 
  ggplot(., aes(name, value, color = key)) + 
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        axis.title.x = element_blank()) +
  scale_y_continuous(name = "Volatility")
```

Mississippi State stands out in terms of volatility. Desipte finishing around 14 in the 2017-2018 season, their preseason rankings were a bit below that. However, they are now beginning to rise in the rankings.

@@

First, we can see which 25 teams have had the most price changes throughout the CFP-RSVP's existence. Note that the following visualizations displays the total number of price changes (includes both upper and lower level changes).

@@

ggplot(changes, aes(name, Count)) + 
  geom_point() + 
  scale_x_discrete(breaks = changes$name) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        axis.title.x = element_blank())
```

With some exceptions (Texas, Florida State, Arizona, Memphis, Miami, Northwestern, and N.C. State), we see that those teams with the most number of changes are well-aligned with the volatility within the teams that have been in the top 25 throughout the season.

We can also plot the reported series for all of those teams which have appeared in the top 25 this season:

```{r}
allHistory %>% 
  filter(allHistory$name %in% top25Ever) %>%
  mutate(name = factor(name, levels = currentRankOrder$name)) %>% 
  group_by(name, variable) %>% 
  mutate(obs = 1:n()) %>% 
  ggplot(., aes(obs, value, color = variable)) + 
  geom_path() +
  facet_wrap(~name) +
  theme_minimal() +
    theme(axis.title.y = element_blank(), 
        axis.title.x = element_blank())
```

Unfortunately, there is no time component to this reported series -- only a series of numbers. So finding out what caused Alabama's dip in the reported numbers would not be possible.

@@

We want to explore if current mean ranking (Mean), wins, injury count (n), and strength of schedule (masseySoS) could be predictive of volatility. Since we want to allow for as much flexibility in our model as possible, we will use generalized additive models for our models. While we will not smooth wins, all other predictors will be smoothed. For the sake of space, we will just look at lower level.

```{r}
predictors = completeData %>% 
  select(Mean, n, name, dateTime, masseyRecord, masseySoS) %>% 
  mutate(wins = as.numeric(stringr::str_extract(masseyRecord, "^[0-9]+"))) %>% 
  group_by(name) %>% 
  slice(n())

@@

modLL = gam(volLL ~ s(Mean) + wins + s(n) + s(masseySoS),
            data = modDat)

@@

knitr::kable(tidy(modLL, parametric = TRUE))
```

For our smoothed terms, we can see that strength of schedule does not do much, but mean ranking and number of injuries do contribute to volatility. Mean ranking behaves as anticipated (a near linear negative relationship), but the number of injuries has an interesting relationship with volatility (see the following plot). Wins, our lone parametric term, did not have much to contribute.

```{r}
par(mfrow = c(2, 2))

plot.gam(modLL, all.terms = TRUE, se = FALSE)
```

@@

We will run a similar model, except using current lower level price as our outcome variable.

```{r}
currentPriceModDat = completeData %>% 
  select(Mean, n, name, dateTime, masseyRecord, masseySoS, UL_price, LL_price) %>% 
  mutate(wins = as.numeric(stringr::str_extract(masseyRecord, "^[0-9]+"))) %>% 
  group_by(name) %>% 
  slice(n())

modLL = gam(LL_price ~ s(Mean) + wins + s(n)+ s(masseySoS),
            data = currentPriceModDat)

@@

knitr::kable(tidy(modLL, parametric = TRUE))
```

While wins and strength of schedule did not offer much (again), mean ranking and number of injuries did offer some predictive power. The effect of mean ranking on the current price took on an interesting pattern (as seen below), while the number of injuries had a negative relationship (as would be expected).

```{r}
par(mfrow = c(2, 2))

plot.gam(modLL, all.terms = TRUE, se = FALSE)
```

## Model Testing

```{r}
currentPriceModDat = na.omit(currentPriceModDat)

set.seed(1001)

trainIndex <- createDataPartition(currentPriceModDat$LL_price, 
                                  p = .50, 
                                  list = FALSE, 
                                  times = 1)

trainDat = currentPriceModDat[ trainIndex, ]

testDat  = currentPriceModDat[-trainIndex, ]

trainMod = gam(UL_price ~ s(Mean) + wins + s(n),
            data = trainDat)

testMod = update(trainMod, data = testDat)
```
